Global tradeoffs between prosperity, inequality, and the environment

### 2023-01-25 paper/manuscript.Rmd
prep new submission, include Tong re-write

---
title: "Global tradeoffs between prosperity, inequality, and the environment"
author: |
    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^, Emilie Lindkvist^2^, Maike Hamann^6^, Jiangxiao Qiu^7^, 
    | \small Caroline Schill^2,8^, Alon Shepon^9^, Anne-Sophie Crépin^8^, and Carl Folke^2,8^
    | \small
    | \footnotesize ^1^The Natural Capital Project and Stanford Center on China's Economy and Institutions, Freeman Spogli 
    | \footnotesize Institute for International Studies, Stanford University
    | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
    | \footnotesize ^3^Future Earth
    | \footnotesize ^4^College of Business and Public Policy, University of Alaska Anchorage
    | \footnotesize ^5^Environment and Sustainability Institute, University of Exeter
    | \footnotesize ^6^Center for Complex Systems in Transition, Stellenbosch University
    | \footnotesize ^7^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
    | \footnotesize ^8^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
    | \footnotesize ^9^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
    | \footnotesize ^§^These authors made equal contributions
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
abstract: |

  Sustainability is often framed as achieving prosperity, equality, and environmental integrity: in other words, high living standards, evenly-distributed, and low-impact. But what if there is a trilemma that inhibits the joint achievement of these three key goals? We analyzed synergies and tradeoffs between these dimensions of sustainability by investigating historical data on per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint. We found that no country in the world successfully solved this trilemma, but there are typologies of countries that can help identify common barriers, and strategies to overcome them. Countries typically reduce inequality and then increase prosperity and environmental impact. Some countries are trapped in a high inequality space, while prosperous countries have reduced inequality only at the expense of increasing environmental impact. Escaping this trilemma to achieve ambitious sustainability goals will likely require economic redistribution and more urgent collective action for planetary stewardship.
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
csl: /Users/juanrocha/Documents/styles/nature.csl
bibliography: references_ei.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
---

@@

@@

Sustainability has often been characterised as having three essential, nested dimensions: economy, society, and the environment. More specifically, sustainability has been defined as attaining high material living standards (prosperity) that are well distributed (equality) and maintainable over the long run (achieved through environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008]. Although each is a desirable goal of development, what if all three are not simultaneously achievable? It has often been assumed that the positive ethical ends to which individuals as well as societies aspire can cohere, but *a priori*, there is no reason to believe that diverse goals, however separately desirable, must be compatible (across scales) or scalable (without upper bounds) [@Gray:2007;
@Greenspan:2001].

In the literature on development, such macro-level incompatibilities are not novel. For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination
[@RodrikD:2011]. Using resources efficiently, fairly and with maximum freedom has also been proposed as an unattainable triad [@Coyle:2011]. Additionally, research suggests that based on international trends, there may be persisting tradeoffs between changes in dietary improvement, disease mitigation, and environmental integrity [@Clark:2018]. In this light, we argue that another such "trilemma" that countries may face in the Anthropocene is the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.

For most of human history, scarcity has been a ubiquitous condition of life: material want, not sufficiency, was the lot of most people [@Maddison:2007;
@Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, approximately 10% of the world’s population still lives under the international poverty line of US\$ 1.90 dollars per day (a derisory standard that the UN itself recognizes may grossly undercount the world's impoverished[@Alston_20220]), with a much larger share barely subsisting above that threshold and aspiring for higher and more secure living standards
[@TheWor:2018]. Furthermore, in a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].

At the same time, while average incomes for many countries in what was once called the "Third World" have progressively converged with those of advanced, mostly Western, economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
The widespread feelings of injustice that such unequal economic distribution arouses has historically been called the "Social Question"[@Faist:2009]. Over the past decade, inequality has become an increasingly urgent issue for policymakers as well as the general public [@Chancel:2020; @UN2020]. This urgency has been strengthened by a growing body of research demonstrating links between inequality and economic complexity, human health and wellbeing, and environmental conditions[@Hamann:2018; @Leach:2018;
@Pikett:2014; @Wilkinson:2009; @Hartmann:2017].

Indeed, environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma [@arrow1995economic]. Human well-being intimately depends on it and it forms the precondition for economic development. There is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere, our home [@Díaz:2019; @folke2021our]. Human activities have become the primary drivers of biodiversity loss, land degradation, and climate change, among other environmental impacts
[@Díaz:2019]. Simultaneously, rapid population and economic growth have exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017]. Global development in the 21st Century requires the protection and restoration of ecosystems as well as the mitigation of catastrophic climate change [@Millenniu:2005; @IPCC:2021].

Recognizing the important nexus formed by economic, social, and environmental challenges, the United Nations’ Sustainable Development Goals explicitly address prosperity, equality, and the environment [@Sachs:2015]. However, while all three of these development goals are necessary, the pursuit of any one may either help or hinder the pursuit of the other two. In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by any given country and by the world as a whole. In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist, are there inherent tradeoffs between these goals, and if it does exist, where does it occur and why does it occur there?

@@

To answer these questions, we examined the historical performance of countries over the past quarter-century using publicly available data on the three dimensions of the sustainability trilemma described above (Table \ref{tab1}). We chose the country scale due to practical considerations
of data availability, but also on the recognition that many of the most
consequential policies affecting sustainability are made and implemented at this level [@Bryan:2018]. We operationalized the trilemma as the movement of countries over time in the three-dimensional space formed by the axes of prosperity, inequality, and environmental footprint  (Fig \ref{fig:trilemma}). There is a development synergy if a country moves in desired directions towards more than one goal (e.g. reducing inequality and increasing prosperity). There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another. Therefore, synergies are movements along
the (blue) diagonal where the achievement of two goals can be positively correlated;
while tradeoffs are represented by the (red) axis where the achievement of one
hinders the achievement of another (negative correlations). It is important to note that in this framing, the sign of the correlation changes depending on the term used, i.e. if one speaks of equality as opposed to inequality, or environmental quality as opposed to environmental footprint. In the remainder of this paper, we use inequality and environmental footprint because that is how the original data are defined and structured. Therefore, across these two domains, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).

@@

To quantify prosperity, equality and environmental impact, we identified a set of widely-used and publicly-available indicators (Table \ref{tab1}). For prosperity, we used
per-capita gross national income (GNI) in US\$ values.GNI is related to gross domestic product (GDP) and the two are close in value for most countries. A key difference is that GNI accounts for earnings from foreign investments. The World Bank uses GNI per capita as its standard measure for the level of development (e.g., low-, medium- or high-income, data source:
<http://data.worldbank.org>). The threshold for being "prosperous" was US\$12,746 per capita per year, which is the World Bank's definition for a high-income country in 2017, the concluding year of our study period [@TheWor:2018].

@@

We used per-capita ecological footprint to measure environmental impact. Ecological footprint is a holistic estimate of the biophysical resources used by human activity, quantified in terms of the land area required for the extraction of materials and energy and the absorption by the environment of pollution and waste [@Wackernagel:1998]. Its unit is the "global hectare", combining measurements of carbon emissions, built-up land, fishing grounds, croplands, forests, and grazing land into an index of environmental impact. In conjunction with measurements of biocapacity – or the extent to which the environment can provide, absorb, and regenerate – the ecological footprint determines whether a given country (or person or city or the world as a whole) is in an unsustainable condition of ecological "deficit". Though it is not exhaustive, the ecological footprint provides the most widely-used and data-rich approximation of overall human impact on the environment, and it accounts for the effects of trade on resource consumption by country. Ecological footprint and biocapacity data were taken from the Global Footprint Network (source: [https://www.footprintnetwork.org/resources/data/](https://www.footprintnetwork.org/resources/data/)). We used 1.68 global hectares per capita as the threshold for environmental impact; any value higher would exceed the global biocapacity (i.e., ecological deficit). In other words, it is the limit at which current generations' needs can be satisfied without compromising the ability of future generations to satisfy their needs.

@@

We used these data to chart the movement of countries across the dimensions of prosperity, equality, and environmental impact over time. We further investigated whether these movements trace out typologies of countries in the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.  We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to test the optimal number of clusters. We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and vector-fitting to show how clusters are pulled apart in the principal component space. All data used are publicly available, and the code used for the analysis is available at: [https://github.com/juanrocha/trilemma](https://github.com/juanrocha/trilemma).

An analysis of the missing values of the publicly available data shows that assessing the change over time in the trilemma spaces was only possible for 53 countries. The most incomplete dataset is the inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values, and inputed them with a linear interpolation. This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).

@@

No country simultaneously achieved high prosperity, low inequality, and low environmental impact. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were strong tradeoffs between inequality and environmental impact (Fig \ref{fig:trilemma}E - as the Gini indicator for income decreases, ecological footprint increases) as well as increasing prosperity and decreasing environmental impact (Fig \ref{fig:trilemma}G- per-capita GNI and ecological footprint move in the same direction). Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden, New Zealand) exhibited low levels of inequality in the past, but show high variation and a trend towards greater income inequality in more recent years.

While these globally-aggregated patterns provide insight into international variations with respect to key dimensions of sustainability, they do not necessarily reveal the diversity of individual country trajectories, nor their typology. To further investigate individual trajectories, we calculated the displacement on the trilemma space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and ecological footprint plane. These two angles and the total distance covered in the three-dimensional space are used to uncover typologies in country trajectories.

@@

There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}). The inequality and ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity and ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality and prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters. Despite this clustering, countries showed distinctive profiles in terms of their direction of movement in the trilemma space (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's Gini and ecological footprint worsened while GNI per capita grew significantly. And while inequality and environmental impact actually fell in Ukraine, prosperity only increased marginally. To better understand country typologies, we then clustered summary statistics (mean and standard deviations) of the three trilemma dimensions, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

The trajectories of countries in the trilemma space from 1995 to 2017 converged into three general types (Fig \ref{fig:pca}A). First, there was a group of countries "trapped" in a region of high inequality, mainly in Latin America. They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$, that is to say, large variations on the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green). The second group showed large displacements within the trilemma space but with no definite direction, characterized by particularly high variations in Gini and per-capita GNI (Fig \ref{fig:pca}A, in yellow). The third group was composed of countries with relatively higher prosperity and ecological footprint; their movements in the trilemma space were relatively well defined, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group). If we ignore temporal dynamics and focus on the mean and standard deviation of the trilemma dimensions as summary statistics, we are able to include more countries (N=140). Consequently, this yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between inequality, prosperity, and environmental impact. The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies, while the second (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia.

@@

Our analysis found that decreasing inequality and ecological impact pulled countries in opposite directions (Fig \ref{fig:pca}) providing strong evidence for the difficulty of achieving both goals at the same time. The historical data shows that countries typically reduce inequality and then increase environmental impact, not the other way around (Fig \ref{fig:cube_ensamble}A). Similarly, countries tend to reduce inequality first, then increase prosperity (Fig \ref{fig:cube_ensamble}B). Importantly, many middle-income countries have levels of inequality similar to high-income countries, but at lower levels of environmental impact and prosperity (Fig \ref{fig:cube_ensamble}B, C). This means that countries can achieve relatively low levels of inequality while attaining and preserving environmental integrity, but our findings show that higher prosperity often leads to higher environmental impact. On the other hand, our results also indicate that some countries struggled to escape areas in the trilemma space characterized by high levels of inequality during the period under consideration. However, it is important to note that past trajectories do not necessarily determine plausible future pathways.

The potential mechanisms of tradeoffs and synergies remain an important research frontier. Notably, tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC). In this framing, the former increases as the latter grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955]. Since this relationship was first posited, there has been conflicting evidence, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020]. This has made the identification of mechanism difficult.  Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or government regulations  [@Pikett:2014]. Our analysis found no evidence for a KC-like relationship, instead finding that increasing inequality tended to precede growing prosperity.

>>>>

---
title: "Global tradeoffs between prosperity, inequality, and the environment"
author: |
    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^, Emilie Lindkvist^2^, Maike Hamann^6^, Jiangxiao Qiu^7^, 
    | \small Caroline Schill^2,8^, Alon Shepon^9^, Anne-Sophie Crépin^8^, and Carl Folke^2,8^
    | \small
    | \footnotesize ^1^The Natural Capital Project and Stanford Center on China's Economy and Institutions, Freeman Spogli 
    | \footnotesize Institute for International Studies, Stanford University
    | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
    | \footnotesize ^3^Future Earth
    | \footnotesize ^4^College of Business and Public Policy, University of Alaska Anchorage
    | \footnotesize ^5^Environment and Sustainability Institute, University of Exeter
    | \footnotesize ^6^Center for Complex Systems in Transition, Stellenbosch University
    | \footnotesize ^7^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
    | \footnotesize ^8^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
    | \footnotesize ^9^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
    | \footnotesize ^§^These authors made equal contributions
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}

output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
csl: /Users/juanrocha/Documents/styles/apa.csl
bibliography: references_ei.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
- \usepackage{setspace}
---

@@

\doublespacing

**Author contributions:**  Conceptualization was a team effort, TW and JCR produced the first manuscript, TW, JCR and JQ analyzed the data. All authors contributed to the writing and editing of the paper.

**Declaration of interests:** None

**Correspondence:** juan.rocha@su.se | +46730747766

# Highlights

- Historical data reveals that no country in the world has reduced inequality and environmental footprint simultanously
- A trilemma exist where countries need to trade-off progress on social and environmental goals
- An inequality trap seems to exist particularly acute in Latinamerica
- Countries trajectories on the trilemma space suggest that in the future social progress can be achieved without compromising the biosphere

\pagebreak
# Acknowledgements

We would like to thank the conveners and other participants of the 2016-2018 Beijer Young Scholars workshops for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support. This work was also supported by funding from Formas (grant number 2018-02318), and the Swedish Research Council Dnr 2018-05862 and Dnr 2018-06139.

# Global tradeoffs between prosperity, inequality, and the environment

\begin{abstract}

  Sustainability has often been defined as jointly achieving prosperity, equality, and environmental integrity: in other words, material living standards that are high, broadly-distributed, and low-impact. But what if there is a “sustainability trilemma” that inhibits the simultaneous achievement of these three developmental goals? We analyzed synergies and tradeoffs between these three key dimensions of sustainability using data on per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017. Over these two decades, we found that no country in the world overcame the hypothesized trilemma. Nonetheless,the trajectories of  countries in the three-dimensional space formed by these key indicators revealed  typologies that identified  common development barriers as well as potential strategies to overcome them. Historically, countries tended to decrease in income inequality before growing in prosperity and environmental impact. Some countries were trapped in a space of high inequality, while prosperous countries lowered inequality only at the expense of increasing environmental impact. Achieving  greater synergy at the nexus of these three sustainability dimensions will likely require countries - especially those that have already become prosperous - to prioritize economic redistribution and environmental stewardship alongside or even before material growth. At the global level, structural inequities in the world economy as well as historical and geographical factors suggest the need for more urgent collective action, especially in the face of the Anthropocene’s worsening risks. Without such concerted efforts, the sustainable development of individual countries, and of the world, may continue to be stymied by this trilemma.

\end{abstract}

@@

Sustainability has often been characterised as having three essential and nested dimensions: economy, society, and the environment. More specifically, this tripartite framework  has been defined as development that attains high material living standards (prosperity) that are well distributed (equality) and maintainable over the long run (environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008]. Although each goal is desirable in its own right, the question nonetheless needs to be asked: What if all three are not simultaneously achievable? It has often been assumed that the positive ends to which individuals as well as societies aspire can cohere, but *a priori*, there is no reason to believe that diverse goals must be compatible (across scales) or scalable (without upper bounds) [@Gray:2007; @Greenspan:2001].

In the literature on development, such macro-level incompatibilities are not novel. For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination [@RodrikD:2011]. Using resources efficiently, fairly and with maximum freedom has also been proposed as an unattainable triad [@Coyle:2011]. Additionally, based on international trends, research suggests that there may be persisting tradeoffs between dietary improvement, disease mitigation, and environmental integrity [@Clark:2018]. In this light, we argue that there may be another such "trilemma" facing countries in the Anthropocene: the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.

For most of human history, material want, not sufficiency, was the norm for most people regardless of county [@Maddison:2007; @Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, a significant share of the world’s population still lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished[@Alston_20220]). Right above this derisory threshold,  hundreds of millions more people maintain a precarious subsistence and understandably aspires for higher and more secure living standards [@TheWor:2018]. Furthermore, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already low poverty line, leading to an increase rise of global poverty unprecedented since the end of World War II and underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].

At the same time, while average incomes for certain countries in what was once called the "Third World" began to  converge with those of the advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
The widespread feelings of injustice that the increasingly unequal economic distribution arouses has historically been called the "Social Question", inspiring influential theories of political transformation in the 19th and 20th Centuries (for instance various strands of Marxist and Third-Worldist thought) [@Christiansen:2019; @Faist:2009]. Over the past decade, inequality has once again become an increasingly urgent issue for policymakers as well as the general public, prompting anxieties about social and political instability [@Chancel:2020; @UN2020]. These anxieties have been reinforced by a growing body of research demonstrating complex linkages between inequality and a range of financial, health, and ecological risks[@Hamann:2018; @Leach:2018; @Pikett:2014; @Wilkinson:2009; @Hartmann:2017]. Furthermore, the impacts and policy responses from the COVID-19 pandemic worsened income and wealth inequalities within many countries, as well as widening the development trajectories between high- and low-income economies [@WorldBank:2022].

Environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma [@arrow1995economic]. Interdisciplinary research in sustainability science confirms that human well-being and economic development depend on the resilience and abundance of nature. At the same time, there is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere, which constitutes the habitative upper bound and common denominator for our species (notwithstanding fanciful visions of extraterrestrial expansion)  [@Díaz:2019; @folke2021our]. Anthropogenic drivers  have become primarily responsible for global biodiversity loss, land degradation, and climate change, among other environmental impacts [@Díaz:2019]. Indeed, humanity  has already exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017]. Global development in the 21st Century therefore urgently requires more sensible stewardship of the biosphere, from ecosystems to the climate [@Millenniu:2005; @IPCC:2021].

The United Nations’ Sustainable Development Goals explicitly recognize the important nexus formed by prosperity, equality, and the environment  [@Sachs:2015]. And in its report charting the path towards a sustainable recovery in the wake of COVID-19, the World Bank also identified the need to achieve “shared prosperity” by reducing inequality and accounting for the climate crisis and other environmental challenges [@WorldBank:2022]. Nonetheless, while all three of these development goals are seen as necessary, it remains for empirical investigation to determine whether the pursuit of any one helps or hinders the pursuit of the other two. In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by any given country and by the world as a whole. In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist and are there inherent tradeoffs between these goals? If a sustainability trilemma does exist, where does it occur and why does it occur there?

@@

To answer these research questions, we examined the performance of countries over the past quarter-century using publicly available data on the three dimensions of the sustainability trilemma described above (Table \ref{tab1}). We chose the country level based on the fact that many of the most consequential policies, plans, and institutions affecting sustainability are made by national governments [@Bryan:2018], and due to practical considerations of data availability. We operationalized the sustainability trilemma as the movement of countries over time in the three-dimensional space formed by the axes of prosperity, (in)equality, and environmental integrity (alternatively, impact)  (Fig \ref{fig:trilemma}). There is a development synergy if a country moves in desired directions towards more than one goal (e.g., reducing inequality and increasing prosperity). There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another. Therefore, synergies between two goals are movements along the (blue) diagonal (positively correlated), while tradeoffs are represented by the (red) axis, where the achievement of one hinders the achievement of another (negatively correlated). It is important to note that in this framing, the sign of the correlation changes depending on the term used. For instance, if one speaks of equality as opposed to inequality, or environmental integrity as opposed to environmental impact. In the remainder of this paper, we use inequality and environmental impact (proxied by ecological footprint) because this is how the original data was defined and structured. Therefore, across these two domains, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).

@@

To quantify prosperity, inequality and environmental impact, we identified a set of widely-used and publicly-available indicators (Table \ref{tab1}). For prosperity, we used per-capita gross national income (GNI) in US\$ values. GNI is related to gross domestic product (GDP) and the two are close in value for most countries. A key difference is that GNI accounts for earnings from foreign investments. The World Bank uses GNI per capita as its standard measure for the level of development (e.g., low-, medium- or high-income, data source: <http://data.worldbank.org>). The threshold for being "prosperous" was US\$12,746 per capita per year, which was the World Bank's definition for a high-income country in 2017, the concluding year of our study period [@TheWor:2018].

@@

We used per-capita ecological footprint to measure environmental impact. Ecological footprint is a holistic estimate of the biophysical resources used by human activity, quantified in terms of the land area required for the extraction of materials and energy and the absorption by the environment of pollution and waste [@Wackernagel:1998]. Its unit is the "global hectare", combining measurements of carbon emissions, built-up land, fishing grounds, croplands, forests, and grazing land into an index of environmental impact. In conjunction with measurements of biocapacity – or the extent to which the environment can provide, absorb, and regenerate – the ecological footprint determines whether a given country (or person or city or the world as a whole) is in an unsustainable condition of ecological "deficit". Though it is not exhaustive, the ecological footprint provides the most widely-used and data-rich approximation of overall human impact on the environment, and it accounts for the effects of trade on resource consumption by country. Ecological footprint and biocapacity data were taken from the Global Footprint Network (source: [https://www.footprintnetwork.org/resources/data/](https://www.footprintnetwork.org/resources/data/)). We used 1.68 global hectares per capita as the threshold for environmental impact; any value higher would exceed global biocapacity (i.e., ecological deficit). In other words, it is the limit at which current generations' needs can be satisfied without compromising the ability of future generations to satisfy their needs.

@@

We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time. We further investigated whether these movements trace out typologies of countries within the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.  We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to determine the relevant number of clusters (clustering is a statistical process whereby objects are partitioned and then grouped in such a way that the members of one cluster are more closely related to each other than they are to those of other clusters). We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and then vector-fitting to show how clusters are pulled apart in the principal component space (in order to make the three-dimensional distribution of countries intelligible). All data used are publicly available, and the code used for the analysis is available at: <link will be added upon publication, avoided here due to double-blind requirements>.

An analysis of the missing values in the publicly available data shows that assessing changes over time in the trilemma space was only possible for 53 countries. The most incomplete dataset is that for inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values and then imputed them with a linear interpolation. This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).

@@

No country simultaneously achieved high prosperity, low inequality, and low environmental impact. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were strong tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreases, ecological footprint increases) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G- per-capita GNI and ecological footprint move in the same direction). Results show synergies between increasing prosperity and reducing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden, New Zealand) exhibited low levels of inequality in the past, but also showed high variation and a trend towards greater income inequality in more recent years.

While these globally-aggregated patterns provide insight into international variations with respect to key dimensions of sustainability, they do not necessarily reveal the diversity of individual country trajectories, nor their typologies. To further investigate individual trajectories, we calculated the displacement within the trilemma space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and ecological footprint plane. These two angles and the total distance covered in the three-dimensional space help to identify typologies in country trajectories.

@@

There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}). The inequality-ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity-ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters. Despite this clustering, countries showed distinctive profiles in terms of their direction of movement (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's Gini and ecological footprint worsened while per-capita GNI grew significantly. And while inequality and environmental impact actually fell in Ukraine, prosperity only increased marginally. To better understand country typologies, we then clustered summary statistics (mean and standard deviations) of the three trilemma dimensions, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

The trajectories of countries in the trilemma space from 1995 to 2017 converged into three general types (Fig \ref{fig:pca}A). First, there was a group of countries "trapped" in a region of high inequality, mainly in Latin America. They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$; that is to say, large variations on the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green). The second group showed large displacements within the trilemma space but with no definite direction, and was characterized by particularly high variations in its Gini coefficient for income and per-capita GNI (Fig \ref{fig:pca}A, in yellow). The third group was composed of countries with relatively higher prosperity and ecological footprint; their movements in the trilemma space were relatively well-defined, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group). If we ignore temporal dynamics and focus on the mean and standard deviation of the trilemma dimensions as summary statistics, we are able to include more countries (N=140). Consequently, this yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between prosperity, inequality, and environmental impact. The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies, while the second (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia.

@@

Overall, our analysis found that decreasing inequality and decreasing environmental impact pulled countries in opposite directions (Fig \ref{fig:pca}), providing strong evidence for the difficulty of achieving both goals at the same time. Also, countries typically decreased in inequality before increasing in environmental impact (Fig \ref{fig:cube_ensamble}A). Similarly, countries tended to see reductions in inequality before prosperity rose (Fig \ref{fig:cube_ensamble}B). Importantly, many middle-income countries had levels of inequality similar to high-income countries, but at lower levels of environmental impact and prosperity (Fig \ref{fig:cube_ensamble}B, C). Overall, our findings suggest that higher prosperity usually comes at the expense of environmental integrity. National trajectories through the three-dimensional trilemma space  also revealed that some countries struggled to escape areas characterized by high inequality, which may represent an “inequality trap” subset of the larger sustainability trilemma.

It is important to note that this study was a historical, not predictive, analysis, and that past trajectories do not necessarily determine plausible future pathways. The potential mechanisms behind these tradeoffs and synergies between prosperity, inequality, and environmental impact constitute an important research frontier in sustainability science and development studies. The patterns and typologies identified in this paper could provide insights or directions for future investigations. Nonetheless, some mechanisms have already been hypothesized for pairwise relationships among the indicators in the often disparate literatures of economics, sociology, and environmental science. In what follows, we review several of the most influential hypotheses in the context of our findings.
